Equivalence Testing and the Second Generation \emph{P}-Value

### 2019-03-09 manuscript.R1.Rmd
Few typos and minor changes

abstract: |
  To move beyond the limitations of null-hypothesis tests, statistical approaches have been developed where the observed data is compared against a range of values that are equivalent to the absence of a meaningful effect. Specifying a range of values around zero allows researchers to statistically reject the presence of effects large enough to matter, and prevents practically insignificant effects from being interpreted as a statistically significant difference. We compare the behavior of the recently proposed second generation *p*-value [@blume_second-generation_2018] with the more established Two One-Sided Tests (TOST) equivalence testing procedure [@schuirmann_comparison_1987]. We show that the two approaches yield almost identical results under optimal conditions. Under suboptimal conditions (e.g., when the confidence interval is wider than the equivalence range, or when confidence intervals are asymmetric) the second generation *p*-value becomes difficult to interpret as a descriptive statistic. The second generation *p*-value is interpretable in a dichotomous manner (i.e., when the SGPV equals 0 or 1 because the confidence intervals lies completely within or outside of the equivalence range), but this dichotomous interpretation does not require calculations. We conclude that equivalence tests yield more consistent *p*-values, distinguish between datasets that yield the same second generation *p*-value, and allow for easier control of Type I and Type II error rates.

@@

To test predictions researchers predominantly rely on null-hypothesis tests. 
This statistical approach can be used to examine whether observed data is sufficiently surprising under the null hypothesis to reject an effect that equals exactly zero. 
Null-hypothesis tests have an important limitation, in that this procedure can only reject the hypothesis that there is no effect, while scientists should also be able to provide statistical support for *equivalence*. 
When testing for equivalence researchers aim to examine whether an observed effect is too small to be considered meaningful, and therefore is practically equivalent to zero. 
By specifying a range around the null hypothesis of values that are deemed practically equivalent to the absence of an effect (i.e., 0 ± 0.3) the observed data can be compared against an *equivalence range* and researchers can test if a meaningful effect is absent [@hauck_new_1984;@kruschke_rejecting_2018;@rogers_using_1993;@serlin_rationality_1985;@spiegelhalter_bayesian_1994;@wellek_testing_2010;@westlake_use_1972].

@@

The TOST procedure also relies on the confidence interval around the effect. 
In the TOST procedure the data is tested against the lower equivalence bound in the first one-sided test, and against the upper equivalence bound in the second one-sided test [@lakens_equivalence_2018]. 
For an excellent discussion of the strengths and weaknesses of different frequentist equivalence tests, including alternatives to the TOST procedure, see [@meyners_equivalence_2012].
If both tests statistically reject an effect as extreme or more extreme than the equivalence bound, you can conclude the observed effect is practically equivalent to zero from a Neyman-Pearson approach to statistical inferences. 
Because one-sided tests are performed, one can also conclude equivalence by checking whether the 1-2$\times$$\alpha$ confidence interval (e.g., when the alpha level is 0.05, a 90% CI) falls completely within the equivalence bounds. 
Because both equivalence tests as the SGPV are based on whether and how much a confidence interval overlaps with equivalence bounds, it seems worthwhile to compare the behavior of the newly proposed SGPV to equivalence tests to examine the unique contribution of the SGPV to the statistical toolbox.

@@

For example, in Figure \ref{fig:TOSTSGPV4} we have plotted four SGPV's. From A to D the SGPV is `r round(SGPV1, digits = 2)`, `r round(SGPV2, digits = 2)`, `r round(SGPV3, digits = 2)`, and `r round(SGPV4, digits = 2)`. The difference in the percentage of overlap between A and B (`r round(SGPV1-SGPV2, digits = 2)`) is identical to the difference in the percentage of overlap between C and D as the mean gets 0.1 closer to the test value (`r round(SGPV3-SGPV4, digits = 2)`).
As the observed mean in a one-sample *t*-test lies closer to the test value, from situation A to D, the mean gets closer to the test value by 0.1) the difference in the overlap changes uniformly. 
As we move the observed mean closer to the test value in steps of 0.1 across A to D the *p*-value calculated for normally distributed data is not uniformly distributed. The probability of observing data more extreme than the upper bound of 2 is (from A to D) `r round(tail1, digits = 3)`, `r round(tail2, digits = 3)`, `r round(tail3, digits = 3)`, and `r round(tail4, digits = 3)`. As we can see, the difference between A and B (`r round(tail1-tail2, digits = 3)`) is not the same as the difference between C and D (`r round(tail3-tail4, digits = 3)`). Indeed, the difference in *p*-values is the largest as you start at *p* = 0.5 (when the observed mean falls on the test value), which is why the line in Figure \ref{fig:TOSTSGPV1} is the steepest at *p* = 0.5. Note that where the SGPV reaches 1 or 0, *p*-values closely approximate 0 and 1, but never reach these values.

@@

##The relation between equivalence tests and SGPV when for asymmetrical confidence intervals around correlations

So far we have only looked at the relation between equivalence tests and the SGPV when confidence intervals are symmetric (e.g., for confidence intervals around mean differences). For correlations, which are bound between -1 and 1, confidence intervals are only symmetric for a correlation of exactly 0. The confidence interval for a correaltion becomes increasingly asymmetric as the observed correlation nears -1 or 1. For example, with ten observations, an observed correlation of 0 has a symmetric 95% confidence interval ranging from -0.630 to 0.630, while and observed correlation of 0.7 has an asymmetric 95% confidence interval ranging from 0.126 to 0.993. Note that calculating confidence intervals for a correlation involves a Fisher's z-transformation, which transforms values such that they are approximately normally distributed, which allows one to compute symmetric confidence intervals, which are then retransformed into a correlation, where the confidence intervals are asymmetric if the correlation is not exactly zero.

@@

This example illustrates the difference between a proportion and a probability. As long as data is normally distributed, there is always a 50% probability of observing a correlation smaller or larger than the true correlation, but the SGPV for this situation depends on how far away the observed correlation is from 0. The further away from 0, the larger the SGPV when the observed mean falls on the equivalence bound. The SGPV is the proportion of values in a 95% confidence interval that overlap with the equivalence range, but not the probability that these values will be observed. In the most extreme case (i.e., a sample size of 4, and equivalence bounds set to r = -0.99 and 0.99, with an observed correlation of 0.99) `r round(sgpv,3)*100`% of the confidence interval overlaps with the equivalence range, even though in the long run only 50% of the correlations observed in the future will fall in this range. It should be noted that in larger sample sizes the SGPV is closer to 0.5 whenever the observed correlation falls on the equivalence bound, but this extreme example nevertheless clearly illustrates the difference between question the SGPV answers, and the question a *p*-value answers. The conclusion of this section on asymmetric confidence intervals is that a SGPV of 1 or 0 can still be interpreted as a *p* < 0.025 or *p* > 0.975 in an equivalence test, since the SGPV and *p*-value for the TOST procedure are always directly related at the values *p* = 0.025 and *p* = 0.975. Although @blume_second-generation_2018 state that "the degree of overlap conveys how compatible the data are with the null premise" this definition of what the SGPV provides does not hold for asymmetric confidence intervals. Although a SGPV of 1 or 0 can be directly interpreted, a SGPV between 0 and 1 is not interpretable as 'compatibility with the null hypothesis' under the assumption of normally distributed data, and the generalizability of this statement needs to be examined beyond normally distributed data. Indeed, Blume and colleagues write in the supplemental material that "The magnitude of an inconclusive second-generation *p*-value can vary slightly when the effect size scale is transformed. However definitive findings, i.e. a *p*-value of 0 or 1 are *not* affected by the scale changes."

@@

When introducing a new statistical method, it is important to compare it to existing approaches and specify its relative strengths and weaknesses. First of all, even though a SGPV of 1 or 0 has a clear interpretation (we can reject effects outside or inside the equivalence range), intermediate values are not as easy to interpret (especially for effects that have asymmetric confidence intervals). In one sense, they are what they are (the proportion of overlap), but it can be unclear what this number tells us about the data we have collected. This is not too problematic, since the main use of the SGPV (e.g., in all examples provided by Blume and colleagues) seems to be to examine whether the SGPV is 0, 1, or inconclusive. As already mentioned, this interpretation of a SGPV as is very similar to the Neyman-Pearson interpretation of an equivalence test and a minimum effect tests (which are complementary). The difference is that where a SGPV of 1 can be interpreted as *p* < .025, equivalence tests provide exact *p*-values, and they continue to differentiate between for example *p* = 0.048 and *p* = 0.002. Whether this is desirable depends on the perspective that is used. From a Neyman-Pearson perspective on statistical inferences the main conclusion is based on whether or not $p < \alpha$, and thus an equivalence test and SGPV can be performed by simply checking whether the confidence interval falls within the equivalence range, just as a null-hypothesis test can be performed by checking whether the confidence interval contains zero or not. At the same time, it is recommended to report exact *p*-values [@american_psychological_association_publication_2010], and exact *p*-values might provide information of interest to readers about how precisely how surprising the data, or more extreme data, is under the null model. Some researchers might be interested in combining an equivalence tests with a null-hypothesis significance test. This allows a researcher to ask whether there is an effect that is statistically different from zero, and whether effect sizes that are considered meaningful can be rejected. Equivalence tests combined with null-hypothesis tests classify results into four possible categories, and for example allow researchers to conclude an effect is significant *and* equivalent (i.e., statistically different from zero, but also too small to be considered meaningful, see [@lakens_equivalence_2018]).

>>>>

abstract: |
  To move beyond the limitations of null-hypothesis tests, statistical approaches have been developed where the observed data are compared against a range of values that are equivalent to the absence of a meaningful effect. Specifying a range of values around zero allows researchers to statistically reject the presence of effects large enough to matter, and prevents practically insignificant effects from being interpreted as a statistically significant difference. We compare the behavior of the recently proposed second generation *p*-value [@blume_second-generation_2018] with the more established Two One-Sided Tests (TOST) equivalence testing procedure [@schuirmann_comparison_1987]. We show that the two approaches yield almost identical results under optimal conditions. Under suboptimal conditions (e.g., when the confidence interval is wider than the equivalence range, or when confidence intervals are asymmetric) the second generation *p*-value becomes difficult to interpret as a descriptive statistic. The second generation *p*-value is interpretable in a dichotomous manner (i.e., when the SGPV equals 0 or 1 because the confidence intervals lies completely within or outside of the equivalence range), but this dichotomous interpretation does not require calculations. We conclude that equivalence tests yield more consistent *p*-values, distinguish between datasets that yield the same second generation *p*-value, and allow for easier control of Type I and Type II error rates.

@@

To test predictions researchers predominantly rely on null-hypothesis tests. 
This statistical approach can be used to examine whether observed data are sufficiently surprising under the null hypothesis to reject an effect that equals exactly zero. 
Null-hypothesis tests have an important limitation, in that this procedure can only reject the hypothesis that there is no effect, while scientists should also be able to provide statistical support for *equivalence*. 
When testing for equivalence researchers aim to examine whether an observed effect is too small to be considered meaningful, and therefore is practically equivalent to zero. 
By specifying a range around the null hypothesis of values that are deemed practically equivalent to the absence of an effect (i.e., 0 ± 0.3) the observed data can be compared against an *equivalence range* and researchers can test if a meaningful effect is absent [@hauck_new_1984;@kruschke_rejecting_2018;@rogers_using_1993;@serlin_rationality_1985;@spiegelhalter_bayesian_1994;@wellek_testing_2010;@westlake_use_1972].

@@

The TOST procedure also relies on the confidence interval around the effect. 
In the TOST procedure the data are tested against the lower equivalence bound in the first one-sided test, and against the upper equivalence bound in the second one-sided test [@lakens_equivalence_2018]. 
For an excellent discussion of the strengths and weaknesses of different frequentist equivalence tests, including alternatives to the TOST procedure, see @meyners_equivalence_2012.
If both tests statistically reject an effect as extreme or more extreme than the equivalence bound, you can conclude the observed effect is practically equivalent to zero from a Neyman-Pearson approach to statistical inferences. 
Because one-sided tests are performed, one can also conclude equivalence by checking whether the 1-2$\times$$\alpha$ confidence interval (e.g., when the alpha level is 0.05, a 90% CI) falls completely within the equivalence bounds. 
Because both equivalence tests as the SGPV are based on whether and how much a confidence interval overlaps with equivalence bounds, it seems worthwhile to compare the behavior of the newly proposed SGPV to equivalence tests to examine the unique contribution of the SGPV to the statistical toolbox.

@@

For example, in Figure \ref{fig:TOSTSGPV4} we have plotted four SGPV's. From A to D the SGPV is `r round(SGPV1, digits = 2)`, `r round(SGPV2, digits = 2)`, `r round(SGPV3, digits = 2)`, and `r round(SGPV4, digits = 2)`. The difference in the percentage of overlap between A and B (`r round(SGPV1-SGPV2, digits = 2)`) is identical to the difference in the percentage of overlap between C and D as the mean gets 0.1 closer to the test value (`r round(SGPV3-SGPV4, digits = 2)`).
As the observed mean in a one-sample *t*-test lies closer to the test value, from situation A to D, the mean gets closer to the test value by 0.1) the difference in the overlap changes uniformly. 
As we move the observed mean closer to the test value in steps of 0.1 across A to D the *p*-value calculated for normally distributed data are not uniformly distributed. The probability of observing data more extreme than the upper bound of 2 is (from A to D) `r round(tail1, digits = 3)`, `r round(tail2, digits = 3)`, `r round(tail3, digits = 3)`, and `r round(tail4, digits = 3)`. As we can see, the difference between A and B (`r round(tail1-tail2, digits = 3)`) is not the same as the difference between C and D (`r round(tail3-tail4, digits = 3)`). Indeed, the difference in *p*-values is the largest as you start at *p* = 0.5 (when the observed mean falls on the test value), which is why the line in Figure \ref{fig:TOSTSGPV1} is the steepest at *p* = 0.5. Note that where the SGPV reaches 1 or 0, *p*-values closely approximate 0 and 1, but never reach these values.

@@

##The relation between equivalence tests and SGPV for asymmetrical confidence intervals around correlations

So far we have only looked at the relation between equivalence tests and the SGPV when confidence intervals are symmetric (e.g., for confidence intervals around mean differences). For correlations, which are bound between -1 and 1, confidence intervals are only symmetric for a correlation of exactly 0. The confidence interval for a correlation becomes increasingly asymmetric as the observed correlation nears -1 or 1. For example, with ten observations, an observed correlation of 0 has a symmetric 95% confidence interval ranging from -0.630 to 0.630, while and observed correlation of 0.7 has an asymmetric 95% confidence interval ranging from 0.126 to 0.993. Note that calculating confidence intervals for a correlation involves a Fisher's z-transformation, which transforms values such that they are approximately normally distributed, which allows one to compute symmetric confidence intervals, which are then retransformed into a correlation, where the confidence intervals are asymmetric if the correlation is not exactly zero.

@@

This example illustrates the difference between a proportion and a probability. As long as data are normally distributed, there is always a 50% probability of observing a correlation smaller or larger than the true correlation, but the SGPV for this situation depends on how far away the observed correlation is from 0. The further away from 0, the larger the SGPV when the observed mean falls on the equivalence bound. The SGPV is the proportion of values in a 95% confidence interval that overlap with the equivalence range, but not the probability that these values will be observed. In the most extreme case (i.e., a sample size of 4, and equivalence bounds set to r = -0.99 and 0.99, with an observed correlation of 0.99) `r round(sgpv,3)*100`% of the confidence interval overlaps with the equivalence range, even though in the long run only 50% of the correlations observed in the future will fall in this range. It should be noted that in larger sample sizes the SGPV is closer to 0.5 whenever the observed correlation falls on the equivalence bound, but this extreme example nevertheless clearly illustrates the difference between question the SGPV answers, and the question a *p*-value answers. The conclusion of this section on asymmetric confidence intervals is that a SGPV of 1 or 0 can still be interpreted as a *p* < 0.025 or *p* > 0.975 in an equivalence test, since the SGPV and *p*-value for the TOST procedure are always directly related at the values *p* = 0.025 and *p* = 0.975. Although @blume_second-generation_2018 state that "the degree of overlap conveys how compatible the data are with the null premise" this definition of what the SGPV provides does not hold for asymmetric confidence intervals. Although a SGPV of 1 or 0 can be directly interpreted, a SGPV between 0 and 1 is not interpretable as 'compatibility with the null hypothesis' under the assumption of normally distributed data, and the generalizability of this statement needs to be examined beyond normally distributed data. Indeed, Blume and colleagues write in the supplemental material that "The magnitude of an inconclusive second-generation *p*-value can vary slightly when the effect size scale is transformed. However definitive findings, i.e. a *p*-value of 0 or 1 are *not* affected by the scale changes."

@@

When introducing a new statistical method, it is important to compare it to existing approaches and specify its relative strengths and weaknesses. First of all, even though a SGPV of 1 or 0 has a clear interpretation (we can reject effects outside or inside the equivalence range), intermediate values are not as easy to interpret (especially for effects that have asymmetric confidence intervals). In one sense, they are what they are (the proportion of overlap), but it can be unclear what this number tells us about the data we have collected. This is not too problematic, since the main use of the SGPV (e.g., in all examples provided by Blume and colleagues) seems to be to examine whether the SGPV is 0, 1, or inconclusive. As already mentioned, this interpretation of a SGPV is very similar to the Neyman-Pearson interpretation of an equivalence test and a minimum effect tests (which are complementary). The difference is that where a SGPV of 1 can be interpreted as *p* < .025, equivalence tests provide exact *p*-values, and they continue to differentiate between for example *p* = 0.048 and *p* = 0.002. Whether this is desirable depends on the perspective that is used. From a Neyman-Pearson perspective on statistical inferences the main conclusion is based on whether or not $p < \alpha$, and thus an equivalence test and SGPV can be performed by simply checking whether the confidence interval falls within the equivalence range, just as a null-hypothesis test can be performed by checking whether the confidence interval contains zero or not. At the same time, it is recommended to report exact *p*-values [@american_psychological_association_publication_2010], and exact *p*-values might provide information of interest to readers about how precisely how surprising the data, or more extreme data, is under the null model. Some researchers might be interested in combining an equivalence tests with a null-hypothesis significance test. This allows a researcher to ask whether there is an effect that is statistically different from zero, and whether effect sizes that are considered meaningful can be rejected. Equivalence tests combined with null-hypothesis tests classify results into four possible categories, and for example allow researchers to conclude an effect is significant *and* equivalent (i.e., statistically different from zero, but also too small to be considered meaningful, see @lakens_equivalence_2018).
